### 2025-08-17 causal-inference.qmd
20250809 - label functions
20250816 - additions

- `X` causes `Y`
- `Y` causes `X`
- `X` and `Y` are bidirectional: `X` causes `Y` and `Y` causes `X`
- a third variable (i.e., [confound](#sec-causalDiagramConfounding)), `Z`, influences both `X` and `Y`
- the association between `X` and `Y` is spurious

@@

In general, to meet the third criterion of causality, one must consider the counterfactual—what would have happened "counter to the fact".
A *counterfactual* is what would have happened in the hypothetical scenario that the cause did not occur [i.e., what would have happened in the absence of the cause; @Shadish2002].
When engaging in causal inference, it is important to consider what would have happened if the hypothetical cause had actually not occurred.
It is like an alternate universe where the cause did not happen.
For instance, consider that we conduct an experiment to randomly assign some players to consume a sports drink before a game and other players to drink only water.
In this case, our treatment/intervention group is the group of players that consumed a sports drink.
The control group is the group players that drank only water.
Now, consider that the players in the treatment group outperform the players in the control group in their football game.
In such a study, we observe what *did happen* when players received a treatment.
The counterfactual is knowledge of what *would have happened* to those same players if they simultaneously had not received treatment [@Shadish2002].
The true causal effect, then, is the difference between what did happen and what would have happened.
However, we cannot observe a counterfactual.
This is the fundamental problem of causal inference—we only observe one potential world/outcome/universe.
That is, we do not know for sure what would have happened to the players who received treatment if those same players had actually not received treatment.
We have a control group, but the control group does not have the same players as the intervention group, and it is impossible for a person to simultaneously receive and not receive treatment.

@@

@@

We cannot directly test or establish causality from a non-experimental research design.
Nevertheless, we can leverage various design features that, in combination with other studies using different research methods, collectively strengthen our ability to make causal inferences.
For instance, there are are no experiments in humans showing that smoking causes cancer—randomly assigning people to smoke or not smoke would not be ethical.
The causal inference that smoking causes cancer was derived from a combination of experimental studies in rodents and observational studies in humans.

@@

@@

Confounding occurs when two variables—that are both caused by another variable(s)—have a spurious or noncausal association [@DOnofrio2020].
That is, two variables share a common cause, and the common cause leads the variables to be associated even though they are not causally related.
The common cause—i.e., the variable that influences the two variables—is known as a confound (or confounder).
An example of confounding is depicted in @fig-counfounding:

>>>>

- `X` causes `Y`
- `Y` causes `X`
- `X` and `Y` are bidirectional: `X` causes `Y` and `Y` causes `X`
- a third variable (i.e., [confound](#sec-causalDiagramConfounding)), `Z`, influences both `X` and `Y`
- the association between `X` and `Y` is spurious (noncausal and due to random chance)

@@

In general, to meet the third criterion of causality, one must consider the counterfactual—what would have happened "counter to the fact".
A *counterfactual* is what would have happened in the hypothetical scenario that the cause did not occur [i.e., what would have happened in the absence of the cause; @Shadish2002].
When engaging in causal inference, it is important to consider what would have happened if the hypothetical cause had actually not occurred.
It is like an alternate universe where the cause did not happen.
For instance, consider that we conduct an experiment to randomly assign some players to consume a sports drink before a game and other players to drink only water.
In this case, our treatment/intervention group is the group of players that consumed a sports drink.
The control group is the group players that drank only water.
Now, consider that the players in the treatment group outperform the players in the control group in their football game.
In such a study, we observe what *did happen* when players received a treatment.
The counterfactual is knowledge of what *would have happened* to those same players if they simultaneously had not received treatment [@Shadish2002].
The true causal effect, then, is the difference between what did happen and what would have happened.
However, we cannot observe a counterfactual.
This is the fundamental problem of causal inference—we only observe one potential world/outcome/universe.
That is, we do not know for sure what would have happened to the players who received treatment if those same players had actually not received treatment.
We have a control group, but the control group does not have the same players as the intervention group, and it is impossible for a person to simultaneously receive and not receive treatment.
Thus, an individual causal effect cannot typically be identified [@DOnofrio2020].

@@

Broadly, approaches for causal inference encompass 1) analysis approaches that account for measured factors and 2) designs that account for unmeasured factors [@DOnofrio2020].

@@

We cannot directly test or establish causality from a non-experimental research design.
Nevertheless, we can leverage various analysis approaches and design features that, in combination with other studies using different research methods, collectively strengthen our ability to make causal inferences.
For instance, there are are no experiments in humans showing that smoking causes cancer—randomly assigning people to smoke or not smoke would not be ethical.
The causal inference that smoking causes cancer was derived from a combination of experimental studies in rodents and observational studies in humans.

@@

Here is an alternative way of specifying the same diagram (more similar to `lavaan` syntax) using the `ggdag::dagify()` and `ggdag::ggdag()` functions:

@@

Confounding occurs when two variables—that are both caused by another variable(s)—have a spurious or noncausal association [@DOnofrio2020].
That is, two variables share a common cause, and the common cause leads the variables to be associated even though they are not causally related.
The common cause—i.e., the variable that influences the two variables—is known as a confound (or confounder).
Nevertheless, causal effects and confounding—though competing hypotheses—are not mutually exclusive [@DOnofrio2020].
A predictor can have an association with an outcome that is partly driven by confounding and partly driven by a causal effect.
A key goal, then is determining the extent to which the observed association reflects confounding versus a causal effect.
An example of confounding is depicted in @fig-counfounding:
